Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-01-23 doc/invt.tex
invt.tex: adding new Appendix A

We wish the biased equilibrium distribution, $\pi_i$,
to be the same as the intended distribution, $p_i$,
which is usually flat,
i.e., $p_i$ constant for all $i$'s,
or nearly flat\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011}.
%
We will generally call such a biased simulation
a flat-distribution-sampling (FDS) simulation.
%
To fulfill this goal, the bias potential must satisfy
%
\begin{equation}
  V_i = \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
according to Eq. \eqref{eq:pi_p_V1}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus furnishes a practical aim of convergence for FDS
methods---adjusting the bias potential to make the
observed distribution, which can be computed from the normalized histogram,
match the intended one.
%
Once this aim is achieved,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.

Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_V1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
However, for an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ with which to start.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
%Another difficulty of this approach is that
Besides, the effectiveness of this approach
depends on the precision of the histogram, $H_i$,
which in turn depends on the simulation length
that must be decided in advance.
%
Thus, the rigidity of this framework disallows
continual improvement of the bias potential, hence that of the PMF,
as simulation lengthens.

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continually and incrementally update the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude decreases according to
the optimal inverse-time prescription\cite{
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

@@

We first approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot V_i(t)
  \equiv
  \frac{ dV_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.

@@

For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_V}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i(t)
  =
  \frac{                p^*_i \, e^{-V_i(t)} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j(t)} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%

\subsubsection{Error and linear approximation}

Next, we need to define the error in the bias potential.
%
This is done by subtracting from the bias potential
two contributions.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
%Since the last two terms of Eq. \eqref{eq:v_def}
%remain constant during the course of simulation,
%the updates to $v_i$'s are
%the same as the updates to $V_i$'s.
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V1},

Second, the bias potential following Eq. \eqref{eq:vt_diffeq}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution $\pi_i(t)$ as given by Eq. \eqref{eq:h_ave},
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
In this way,
the variables, $x_i(t)$'s,
represent the error of the bias potential,
and are expected to vanish on convergence.
%

Note that the distribution, $\pmb \rho$,
introduced above
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%
The average growth rate of $v(t)$ is given by
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%

In the asymptotic regime,
we expect the deviation $x_i(t)$
to be small.
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{
The second step of Eq. \eqref{eq:hdet_linearize} follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
The expression depends implicitly on the function $\alpha(t)$.
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
It suffices to consider
a one-variable problem that
minimizes each individual $\left\langle x_i^2 \right\rangle$.

@@

We now consider a multiple-bin updating scheme,
given in Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential at bin $j$,
but also those at a few neighboring bins, $i$'s.
%
The relative updating magnitudes,
$w_{ij}$'s,
form a matrix $\mathbf w$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \sum_{j=1}^n \, w_{ij} \frac{ h_j(t) } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
 $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eqs. \eqref{eq:vbar_def}
and \eqref{eq:w_balance},
we can show that Eq. \eqref{eq:dvbardt}
remains true,
and by using Eqs.
\eqref{eq:x_def},
\eqref{eq:rhox_sum},
\eqref{eq:hdet_linearize},
\eqref{eq:xi_def},
and \eqref{eq:w_sumj}, we get
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left(
    \frac{ h_j(t) } { p_j }
    -
    \sum_{r = 1}^n
      \rho_r \frac{ h_r(t) } { p_r }
  \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \left[ x_j(t) - \xi_j (t) \right]
  .
\end{aligned}
$$
%where
%we have expanded the right-hand side
%in linear terms of $x_j$
%on the second line.

@@

Our method of analysis is most useful for long simulations
to achieve relatively precise results.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one\cite{barducci2008, dama2014},
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although these assumptions appeared to work well
on our test system,
they may break down for relatively short simulations
of glassy systems.
%Our immediate interest is to apply these relations to
%problems containing aqueous mixtures of proteins and nucleic acids.

>>>>

We wish the biased equilibrium distribution, $\pi_i$,
to be the same as the intended distribution, $p_i$,
which is usually flat,
i.e., $p_i$ constant for all $i$'s,
or nearly flat\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011}.
%
We will generally call such a biased simulation
a flat-distribution-sampling (FDS) simulation.
%
To fulfill this goal, the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
according to Eq. \eqref{eq:pi_p_V1}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus sets a practical goal for the convergence of FDS
methods---adjusting the bias potential to make the
observed distribution, or the normalized histogram,
match the intended one.
%
Once this aim is achieved,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering\cite{zhang2007},
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential $V_i$ is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the normalized histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_V1}
allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
However, for an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ with which to start.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
%Another difficulty of this approach is that
Besides, the effectiveness of this approach
depends on the precision of the histogram, $H_i$,
which in turn depends on the simulation length
that must be decided in advance.
%
Thus, the rigidity of this framework disallows
continuous improvement of the bias potential, hence that of the PMF,
as simulation lengthens.

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continuously and incrementally update the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude decreases according to
the optimal inverse-time prescription\cite{
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

@@

\subsubsection{Error}

First, we need to define the error in the bias potential.
%
This is done by subtracting from the bias potential
two contributions.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
the same as the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V1},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
In this way,
the variables, $x_i(t)$'s,
represent the error of the bias potential,
and are expected to vanish on convergence.
%
We, therefore, define
the error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}

Note that the distribution, $\pmb \rho$,
introduced above is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot V_i(t)
  \equiv
  \frac{ dV_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
$
h_i(t) = \delta_{i, i(t)}
$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.

@@

For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_V}
for the deterministic part
%
\begin{align}
  \langle h_i(t) \rangle
  &\approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-x_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-x_j(t)} }
  \notag
  \\
  &\approx
  p_i \,
  \left(
    1 - x_i + \sum_{r=1}^n p_r \, x_r
  \right)
  ,
  \label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $x_i(t)$ in the linear expansion
and used $\sum_{j=1}^n p_j = 1$
on the second line.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%

The average growth rate of $v(t)$ is given by
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%

From Eqs.
\eqref{eq:x_def},
\eqref{eq:rhox_sum}-\eqref{eq:dvbardt},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:h_ave}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error defined in Eq. \eqref{eq:error_sum}
now depends implicitly on the function $\alpha(t)$
via $x_i(t)$.
%
Our aim is to find the $\alpha(t)$
that minimizes the error
at the end of the simulation period $T$.
%
It suffices to consider
a one-variable problem that
minimizes each individual $\left\langle x_i^2 \right\rangle$.

@@

We now consider a multiple-bin updating scheme,
given in Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential at bin $j$,
but also those at a few neighboring bins, $i$'s.
%
The relative updating magnitudes,
$w_{ij}$'s,
form a matrix $\mathbf w$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
%
\begin{equation}
  \dot V_i(t)
  =
  \alpha(t) \sum_{j=1}^n \, w_{ij} \frac{ h_j(t) } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
 $\dot V_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows the bias potential to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eqs. \eqref{eq:vbar_def}
and \eqref{eq:w_balance},
we can show that Eq. \eqref{eq:dvbardt}
remains true,
and by using Eqs.
\eqref{eq:x_def},
\eqref{eq:rhox_sum},
\eqref{eq:h_ave},
\eqref{eq:xi_def},
and \eqref{eq:w_sumj}, we get
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left(
    \frac{ h_j(t) } { p_j }
    -
    \sum_{r = 1}^n
      \rho_r \frac{ h_r(t) } { p_r }
  \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \left[ x_j(t) - \xi_j (t) \right]
  .
\end{aligned}
$$
%where
%we have expanded the right-hand side
%in linear terms of $x_j$
%on the second line.

@@

In this analysis, we have ignored the \emph{systematic} error\cite{zhou2005}
caused by the updates in adaptive FDS methods.
%
This is a valid assumption in the asymptotic regime,
where the random error is much greater than the systematic one.
%
%Our method of analysis is most useful for long simulations
%to achieve relatively precise results.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one\cite{barducci2008, dama2014},
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although this assumption appeared to work well
on our test system for modeling long simulations,
they may break down for relatively short simulations
of glassy systems.
%Our immediate interest is to apply these relations to
%problems containing aqueous mixtures of proteins and nucleic acids.
